Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-26 analysis/paper/paper.Rmd

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail the capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which is usually unavailable [@Cramer2009].

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance (and density) and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

In this study, we developed a QRF model to

@@

This manuscript incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.

@@

Habitat data used in our study were collected from `r as.character(english(n_distinct(hab_data$Watershed)))` watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The Columbia River basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds (see Table \@ref(tab:fish-hab-sites) and Figure \@ref(fig:wtsd-map)), at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin. The sites were selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

@@

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e. not highly correlated), described many aspects of stream habitat (e.g. substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided. Full details of how covariates used in the QRF model were selected can be found in Appendix S1.

@@

After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those within the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e. fall emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick [@Froese2008], using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

All code and data for the analyses presented here can be found on a [GitHub repository](https://github.com/KevinSee/QRFpaper).

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars) and the summary of the model fit is shown in Table \@ref(tab:extrap-r2-table). From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, at roughly a one kilometer scale.

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

The favorable comparison between QRF estimates of watershed capacity and the spawner-recruit estimates in selected watersheds helps support and validate this approach. Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.

>>>>

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods often fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat rehabilitation can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*O. mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for several salmonid species. @Nickelson1992 demonstrated that juvenile coho salmon (*O. kisutch*) were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye salmon (*O. nerka*) in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook salmon (*O. tshawytscha*) in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have used other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout (*O. clarkii*) densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole (*Solea solea*) in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships and provide motivation for furthering research in this realm..

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon (*Salmo salar*) parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners, data which are usually unavailable [@Cramer2009].

For the purposes of this paper, we define carrying capacity as the maximum number of individuals that can be supported given the quantity and quality of habitat available at a given life-stage. We assume that higher observed relative densities within a given life stage are a function of habitat quantity and quality. Furthermore, we assert that observed fish density is a poor predictor of habitat capacity owing to both a paucity of individuals, especially for threatened or endangered species, and the existence of unmeasured variables that may serve to limit capacity. To address this, we have developed a model to estimate juvenile rearing capacity for Pacific salmon in wadeable streams based on quantile random forest (QRF) [@Meinshausen2006] models using measurements of fish abundance and density and habitat characteristics. QRF models combine the theory and justification of quantile regression modeling [@Koenker1978; @Cade2003] with the flexibility and framework of random forest models [@Breiman2001]. They account for unmeasured variables and can be used to describe the entire distribution of predicted fish densities for a given set of habitat conditions, not just the mean expected density. Random forest models have been shown to outperform more standard parametric models in predicting fish-habitat relationships in other contexts [@Knudby2010]. Quantile regression forests share many of the benefits of random forest models, such as the ability to capture non-linear relationships between independent and dependent variables, naturally incorporate interactions between covariates, and work with untransformed data while being robust to outliers [@Prasad2006]. Meanwhile, quantile regression models have been used in a variety of ecological systems to estimate the effect of limiting factors [@Terrell1996; @Cade2003].

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon adults enter freshwater from the ocean earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in freshwater before entering the ocean as smolts. Ocean-type Chinook salmon adults enter freshwater later (e.g., fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean as subyearlings. Here, we focus on stream-type Chinook salmon, and in particular the juvenile summer rearing period during low flow, during which juveniles are often termed parr, referring to the camouflage markings that occur on their sides during this life-stage. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

In this study, we developed a QRF model to:

@@

Our study incorporates multiple measures of stream habitat to estimate fish-habitat relationships that encompass the collinear nature of most stream habitat metrics and can be used to predict carrying capacity. Our approach moves across several spatial scales, inferring fish-habitat relationships from detailed, localized habitat data and extrapolating capacity predictions across wide swaths of unsampled locations.

@@

Habitat data used in our study were collected from `r as.character(english(n_distinct(hab_data$Watershed)))` watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The Columbia River basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds (see Table \@ref(tab:fish-hab-sites) and Figure \@ref(fig:wtsd-map)), at CHaMP survey reaches and were graciously provided by a number of agencies and projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 600 m reaches within wadeable streams across select watershed within the interior Columbia River basin. The sites were selected based on a spatially balanced generalized random tesselation stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, stream size (depth, width, discharge), substrate, temperature, macroinvertebrate productivity, and water quality.

@@

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e., not highly correlated), described many aspects of stream habitat (e.g., substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumably because they contained information about what types of habitat fish sought or avoided. Full details of how covariates used in the QRF model were selected can be found in Appendix S1.

@@

After model fitting, the QRF model was then used to predict capacity at sites with measurements of the habitat covariates that were used to fit the model. In our case, this includes all sites within CHaMP basins in the interior Columbia River basin. For CHaMP sites that were sampled in multiple years, we first calculated the mean for each habitat metric among years to make predictions. In total, we generated `r filter(all_preds, is.na(chnk_per_m_se)) %>% summarise(n_sites = n_distinct(Site)) %>% pull(n_sites)` predictions of Chinook salmon parr capacity, during summer months, for the following basins: Entiat, Grande Ronde (including Minam), John Day, Lemhi, Methow, South Fork Salmon, Tucannon, Wenatchee and Yankee Fork Salmon.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed an extrapolation model based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of generalized random tesselation stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, elevation, watershed, the first principal component of a natural feature classification and human disturbance classificaiton [@Whittier2011], the square root of cumulative drainage area, stream power, slope, channel type, bankfull width and average August temperature (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions (parr/m) was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two different extrapolation models, one that included watershed as a covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart.

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those within the domain of spring/summer-run Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org)) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e., fall emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick [@Froese2008], using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

All code and data for the analyses presented here can be found in a [GitHub repository](https://github.com/KevinSee/QRFpaper).

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars) and the summary of the model fit in Table \@ref(tab:extrap-r2-table). From this, we calculated estimates of capacity at every master sample point in the Columbia River basin, each representing roughly one kilometer of stream length.

@@

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months and in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions and help guide not only the type of action, but also the location at which an action is performed.

The favorable comparison between QRF estimates of carrying capacity and the spawner-recruit based estimates in select watersheds helps support and validate this approach. Although built from completely different data, when these multiple lines of evidence converge it lends credence to the QRF capacity prediction results.
